Agda Meets Accelerate

### 2013-03-10 paper/main.tex
Few more changes

As \texttt{Shape} is an ordinary data type in Agda, we can define a
\texttt{size} function that computes the number of elements stored in
an array of a certain shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.
\begin{alltt}
reshape : \{sh : Shape\} \{E : Elt\}
       -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
       -> Array sh' E
\end{alltt}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument.

Furthermore, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate implementation
of \texttt{zipWith}. Instead of checking the sizes of the input
arrays, it truncates them to the respective minima. We also developed
an Agda type that directly corresponds to this implementation. It
requires a binary function \texttt{isect} that computes the minimum of
two shapes of the same rank,
which we leave as an exercise to the reader.
\begin{alltt}
zipWith' : \{A B C\} \{shA shB\} \{p : rank shA \(\equiv\) rank shB\}
         -> (A -> B -> C)
         -> Array shA A -> Array shB B -> Array (isect shA shB p) C
\end{alltt}
\subsection{Associativity of Operations}
\label{sec:assoc-oper}

@@

Because of Agda's lack of support for overloaded numeric literals,
we embed numeric literals for integers and floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Elt}.
The \texttt{:::} operation is the workhorse of the embedding:
\begin{verbatim}
_:::_ : (s : String) -> (E : Elt) 
     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are inferred by Agda.
As mentioned before, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value in scope
\cite{DevriesePiessens2011}. As before, the predicate 
\texttt{Numeric} plays the role of a type class that characterizes
the numeric types.

The function \texttt{parsesAs} dispatches on its ``type'' argument and
parses the string to check whether it is an acceptable literal of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries,
which are computed from \texttt{E} using the functions \texttt{EltDict}
and \texttt{ReadDict}.

This setup results in a flexible way of handling literals. It has
worked well in our examples.

@@

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : forall {n m : Nat}{E : Elt}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result
cannot be determined statically. In fact, the only thing that we could
prove from an implementation of \texttt{filter} is that
\texttt{m} must be less than or equal to \texttt{n}.
However, we cannot even prove this restriction on \texttt{m} for two
reasons. First, the code of \texttt{filter} is not available as even
the underlying Haskell library only deals with abstract
syntax. Second, even if the code were available, there is still a
problem due to staging. The Accelerate implementation computes the length of the result only when the generated GPU code is executed. The function \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax; it does not directly implement a Boolean predicate. Hence, we cannot use the predicate in the result type of \texttt{filter} to more accurately constrain the size of the resulting vector --- unless we include an evaluator for Accelerate expressions.

@@

However, an alternative encoding of arrays can be
used which is compatible with filtering of elements. The idea of this
encoding is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to pair up each element with a boolean flag that indicates
its presence, which we call \emph{predicated arrays}:\footnote{Using an array of \texttt{Maybe}-typed elements is not an option
  because Accelerate does not support \texttt{Maybe} types as array elements.}
\begin{verbatim}
FVector : Nat -> Elt -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}

In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although the result might
require careful interpretation.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool E)
        g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
The mapping operation that applies a function to each element of an
array becomes more complicated because it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool F)
        g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                           else (pair (fst bx) defaultF)
\end{verbatim}
Some operations can get rid of absent
elements. A fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Operations like \texttt{fold1} and the scan operations
extend to this representation, but they
% retain a notion of absent elements and
cannot revert to a
non-filtered representation.

@@

Alternatively, an interactively developed program may be compiled to
Haskell using the Alonzo compiler. This compiler supports a Haskell
foreign function interface (FFI), which enables Agda programs to invoke
Haskell functions.

Using this interface amounts to declaring a typed identifier in Agda
and then binding the identifier to a suitably typed Haskell
function. As an example, consider the import of the \texttt{use}
function.
\begin{code}
postulate 
  useHs : {E : Set}
      -> HsEltDict E -> HsArray HsDIM1 E -> Acc (AccArray HsDIM1 E)
  {-# COMPILED useHs       (\ _ -> Accel.use) #-}
\end{code}
The first three lines introduce the typed identifier \texttt{useHs}
and the last line is a pragma for the Alonzo compiler that binds the
Agda identifier \texttt{useHs} to the Haskell expression on the right.
But wait, this type looks very unpleasant and quite different to the
one mentioned in Section~\ref{sec:array-types}. This difference arises
because the type translation of Alonzo is unable to cope with the
index type \texttt{Shape}. For that reason, the interface uses a
simplified array type and adapter functions are required, in the worst
case, both on the Agda side and on the Haskell side of the interface.

>>>>

As \texttt{Shape} is an ordinary data type in Agda, we can define a
\texttt{size} function that computes the number of elements stored in
an array of a certain shape.
\begin{verbatim}
size : Shape -> Nat
size Z = 1
size (sh :< n >) = size sh * n
\end{verbatim}
Now we can state an accurate type for \texttt{reshape} in Agda, which
involves an extra argument with a proof that the source and
target shapes have the same size.
\begin{alltt}
reshape : \{sh : Shape\} \{E : Elt\}
       -> (sh' : Shape) -> Array sh E -> (size sh \(\equiv\) size sh')
       -> Array sh' E
\end{alltt}
There is a subtle difference to the original signature. In
Accelerate, the first argument is an \emph{expression} that produces a
value of type \texttt{sh} at run time, whereas the Agda
\texttt{reshape} requires a \texttt{Shape} as its first argument. Hence, 
Agda \texttt{reshape} computes the shape in the host language on the CPU,
whereas the original signature admits to compute the new shape in the
embedded language as part of a GPU computation. In other words, we slightly
restrict expressiveness here to gain more static information, we will get
back to that issue when discussing filtering.

In Agda, functions like \texttt{map} and \texttt{zipWith} obtain
more precise types. The type of \texttt{map} tells us that the input
shape is identical to the output shape:
\begin{verbatim}
map : {A B} {sh} -> (Exp A -> Exp B) -> Array sh A -> Array sh B
\end{verbatim}
Similarly, the type of \texttt{zipWith} restricts its input arrays to
identical shapes:
\begin{verbatim}
zipWith : {A B C} {sh} -> (Exp A -> Exp B -> Exp C)
        -> Array sh A -> Array sh B -> Array sh C
\end{verbatim}
The latter type is more restrictive than the Accelerate implementation
of \texttt{zipWith}. Instead of checking the sizes of the input
arrays, it truncates them to the respective minima. We also developed
an Agda type that directly corresponds to this implementation. It
requires a binary function \texttt{isect} that computes the minimum of
two shapes of the same rank,
which we leave as an exercise to the reader.
\begin{alltt}
zipWith' : \{A B C\} \{shA shB\} \{p : rank shA \(\equiv\) rank shB\}
         -> (A -> B -> C)
         -> Array shA A -> Array shB B -> Array (isect shA shB p) C
\end{alltt}
\subsection{Associativity of Operations}
\label{sec:assoc-oper}

@@

Because of Agda's lack of support for overloaded numeric literals,
we embed numeric literals for integers and floating point numbers
using a string with an explicit type annotation that determines the
parsing of the string. Here are some example embeddings:
\begin{verbatim}
"3.1415926" ::: Float
"6.0221415E23" ::: Double
\end{verbatim}
Recall that \texttt{Float} and \texttt{Double} are not
types, but rather values of type \texttt{Elt}.
The \texttt{:::} operation is the workhorse of the embedding:
\begin{verbatim}
_:::_ : (s : String) -> (E : Elt) 
     -> {{nu : Numeric E}} -> {p : T (s parsesAs E)} -> Exp E
s ::: E = Ex (constantFromString (EltDict E) (ReadDict E) s)
\end{verbatim}
The arguments \texttt{s} and \texttt{E} are explicit, but the
remaining ones are inferred by Agda.
As mentioned, the argument \texttt{nu} is an instance argument; it is automatically
filled-in with a suitably typed value in scope
\cite{DevriesePiessens2011}. As before, the predicate 
\texttt{Numeric} plays the role of a type class that characterizes
the numeric types.

The function \texttt{parsesAs} dispatches on its ``type'' argument and
parses the string to check whether it is an acceptable literal of the expected type. The
function \texttt{constantFromString} is imported from Accelerate.
It is an overloaded function that requires two type dictionaries,
which are computed from \texttt{E} using the functions \texttt{EltDict}
and \texttt{ReadDict}. 
This results in a flexible way of handling literals, which
worked well in our examples.

@@

For a related problem, consider an implementation of the \texttt{filter}
operation that takes a predicate and a source array and returns an
array that only contains the elements of the source array fulfilling
the predicate.  First of all, filtering only makes sense for
one-dimensional arrays, that is, for vectors. To see the second catch,
let's try to write down a dependent type signature for \texttt{filter}.
\begin{verbatim}
filter : forall {n m : Nat}{E : Elt}
       -> Vector n E -> (Exp E -> Exp Bool) -> Vector m E
\end{verbatim}
The problem is that the size of the result cannot be determined statically --- that is,
we cannot simplify define a type-level function that determines the length of a vector.
Why? \texttt{Vector m E} is not a representation of a vector. Instead, it is a representation
of a computation that, \emph{once run,} produces a vector.

Similarly, we cannot define a function that uses the predicate passed to \texttt{filter} to 
count the number of elements that will appear in \texttt{filter}'s result. Such a function would
need access to the elements of the \texttt{filter}ed vector, but, as discussed, we cannot even get
its length. Moreover, such a counting function would need to evaluate the predicate. We cannot do
that as the predicate of type \texttt{Exp E -> Exp Bool} maps abstract syntax to abstract syntax;
it \emph{does not} directly implement a Boolean predicate. We might consider to include an evaluator
for abstract syntax to lift these restrictions. However, that evaluator would not be the code
actually executed on the GPU, and hence, it doesn't seem to be any more valuable than simply
asserting an axiom concerning the size of \texttt{filter}'s result.
That is the price we pay for a generative approach, where at program runtime, we dynamically
generate the code to be executed on the GPU.

We encounter similar restrictions if we try to, at least, establish that \texttt{m} must be less
than or equal to \texttt{n} for \texttt{filter}. We cannot prove this constraint as the GPU code of
\texttt{filter} is not available to us --- it is generated by the underlying Haskell library. Even
if we had access to that code, any statements about its properties would need to be based on the
semantics of CUDA (i.e., NVIDIA's C dialect for GPU programming).

@@

However, we may use an alternative encoding of arrays that is compatible with filtering. The idea
is to keep all elements but mark those which are no longer
present because they have been filtered out.
There are several ways of implementing this idea. The simplest
approach is to pair up each element with a boolean flag that indicates
its presence, which we call \emph{predicated arrays}:\footnote{Accelerate currently does not support
\texttt{Maybe} types as array elements.}
%
\begin{verbatim}
FVector : Nat -> Elt -> Set
FVector n E = Vector n (Pair Bool E)
\end{verbatim}
%
In this encoding, filtering is quite simple because the length of
the \texttt{FVector} does not change. Furthermore, filtering could be
extended to multi-dimensional arrays, although the result might
require careful interpretation.
\begin{verbatim}
filterF : forall {n : Nat}{E : Elt}
        -> (Exp E -> Exp Bool) -> FVector n E -> FVector n E
filterF {n}{E} pred vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool E)
        g bx = pair ((fst bx) && p (snd bx)) x
\end{verbatim}
Mapping, which applies a function to each element of an
array, becomes more complicated as it either has to 
materialize a dummy result for each absent element in the argument
vector or apply the function to absent elements, too. This makes \texttt{filter} reminiscent of the \texttt{where} statement of the SIMD language C$^*$~\cite{rose-etal:c-star}.
\begin{verbatim}
mapF : forall {n : Nat}{E F : Elt}
     -> Exp F -> (Exp E -> Exp F) -> FVector n E -> FVector n F
mapF {n}{E}{F} defaultF f vec = map g vec
  where g : Exp (Pair Bool E) -> Exp (Pair Bool F)
        g bx = if (fst bx) then (pair (fst bx) (f (snd bx)))
                           else (pair (fst bx) defaultF)
\end{verbatim}
Some operations can get rid of absent
elements. A fold operation which reduces a filtered
vector with a monoid returns a single value. In Accelerate, such a value has type
\texttt{Scalar}, which is a synonym for an array of dimension $0$.
\begin{verbatim}
foldF : forall {n : Nat}{E : Elt}
      -> (Exp E -> Exp E -> Exp E) -> Exp E
      -> FVector n E -> Scalar E
foldF f e vec =
  fold f e (map (\ bx -> if (fst bx) then (snd bx) else e) vec)
\end{verbatim}
Operations like \texttt{fold1} and the scan operations
extend to this representation, but they
% retain a notion of absent elements and
cannot revert to a
non-filtered representation.

@@

Alternatively, an interactively developed program may be compiled to
Haskell using the Alonzo compiler. It supports a Haskell
foreign function interface (FFI), for Agda programs to invoke
Haskell functions. 
Using this interface amounts to declaring a typed identifier in Agda
and then binding the identifier to a suitably typed Haskell
function. As an example, consider the import of the \texttt{use}
function.
\begin{code}
postulate 
  useHs : {E : Set}
      -> HsEltDict E -> HsArray HsDIM1 E -> Acc (AccArray HsDIM1 E)
  {-# COMPILED useHs       (\ _ -> Accel.use) #-}
\end{code}
The first three lines introduce the typed identifier \texttt{useHs}
and the last line is a pragma for the Alonzo compiler that binds the
Agda identifier \texttt{useHs} to the Haskell expression on the right.
But wait, this type looks very unpleasant and quite different to the
one mentioned in Section~\ref{sec:array-types}. This difference arises
as the type translation of Alonzo is unable to cope with the
index type \texttt{Shape}. Hence, the interface uses a
simplified array type and adapter functions are required, in the worst
case, both on the Agda side and on the Haskell side of the interface.
